Efficient pedigree recording for fast population genetics simulation

### 2018-06-07 forwards_paper.tex
jerome fixes

\paragraph{Essential steps in tree sequence recording}
Since a tree sequence can record the history of genetic inheritance in any situation
(requiring only unambiguous inheritance and no time travel),
any individual-based population genetics simulator can maintain a tree sequence
with only a little bookkeeping. \revpoint{1}{1}
We have furthermore provided several tools to minimize this bookkeeping,
only requiring one-way output at the birth of each new individual.
Concretely, to record a tree sequence, including mutations,
a simulator must record for each new genome
(so, twice for each new diploid individual): \revpoint{1}{6}
\begin{itemize}
    \item the birth time of the genome in the Node Table,
    \item the segments the genome inherits from its parental genomes in the Edge Table,
    \item the locations of any new mutations in the Site Table,
    \item and the derived state of these new mutations in the Mutation Table
        (as well as the identiy of this genome the mutations appeared in).
\end{itemize}
Each of these can be simply appended to the ends of the respective tables.
Besides this, simplification should be run every once in a while (e.g., every 100 generations).
Before simplification, time in the Node Table must be translated to ``time ago'' if it is not already.
(This was avoided in Algorithm~\algref{W} since there $t$ denoted
``time until the end of the simulation''.)
There are also several ``cleaning'' steps, for which we provide functions in the Tables API:
\emph{sorting} according to several criteria for algorithmic efficiency;
and removing any duplicate sites from the Site Table.
After simplification, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
being relabeled in \tskit{}'s tables as $0, 1, \ldots, N-1$,
there must be bookkeeping that keeps in sync the individual IDs as recorded by the simulator
with the node IDs recorded in the tables.

@@

%% simplify run time
How does the computation \emph{time} required for simplification scale?
Simply because it must process each edge,
% below per discussion
% https://github.com/petrelharp/ftprime_ms/pull/37#discussion_r156617283
the simplification algorithm is at least linear in the number of edges of the input tree sequence.
Empirically the algorithm is exactly linear,
as seen in Figure~\ref{fig:simplify_complexity}B,
which shows the time required to simplify increasingly large subsets of a large tree sequence.
When simplifying the result of a forwards-time sequence, the number of edges is the main contributing factor.
Suppose on the other hand we want to
simplify an already-minimal but large tree sequence with $N$ nodes
to a subsample of size $n$.
How does the required time scale with $n$?
In this case, the computation is dominated by the size of the output tree sequence,
which grows with $\log(n)$, as shown in Figure~\ref{fig:simplify_complexity}C,

@@

In this article, we applied our methods for storing trees to the problem of pedigree recording in a forward-time
simulation.  However, the method applies to any simulation scheme generating nodes and edges.  For example, one could
use the methods described here to generate succinct tree sequences under coalescent processes not currently implemented
in \msprime{}, such as the coalescent with gene conversion \citep{Wiuf2000-rc}, using the structured coalescent to
model various forms of natural selection \citep{Kaplan1988-in,Kaplan1989-rt,Braverman1995-gn},
or the coalescent within a known pedigree.  For such models, one
could in principle generate tables of nodes and edges to be simplified in \tskit{}.  The resulting succinct tree
sequence object would be in the same format as those generated by \tskit{}'s built-in simulate function, and therefore
compatible with existing methods for downstream analyses.

>>>>

\paragraph{Essential steps in tree sequence recording}
Since a tree sequence can record the history of genetic inheritance in any situation
(requiring only unambiguous inheritance and no time travel),
any individual-based population genetics simulator can maintain a tree sequence
with only a little bookkeeping. \revpoint{1}{1}
We have furthermore provided several tools to minimize this bookkeeping,
only requiring one-way output at the birth of each new individual.
Concretely, to record a tree sequence, including mutations,
a simulator must record for each new genome
(so, twice for each new diploid individual): \revpoint{1}{6}
\begin{itemize}
    \item the birth time of the genome in the Node Table,
    \item the segments the genome inherits from its parental genomes in the Edge Table,
    \item the locations of any new mutations in the Site Table,
    \item and the derived state of these new mutations in the Mutation Table
        (as well as the identity of this genome the mutations appeared in).
\end{itemize}
Each of these can be simply appended to the ends of the respective tables.
Besides this, simplification should be run every once in a while (e.g., every 100 generations).
Before simplification, time in the Node Table must be translated to ``time ago'' if it is not already.
(This was avoided in Algorithm~\algref{W} since there $t$ denoted
``time until the end of the simulation''.)
There are also several ``cleaning'' steps, for which we provide functions in the Tables API:
\emph{sorting} according to several criteria for algorithmic efficiency;
and removing any duplicate sites from the Site Table.
After simplification, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
being relabeled in \tskit{}'s tables as $0, 1, \ldots, N-1$,
there must be bookkeeping that keeps in sync the individual IDs as recorded by the simulator
with the node IDs recorded in the tables.

@@

%% simplify run time
How does the computation \emph{time} required for simplification scale?
Simply because it must process each edge,
% below per discussion
% https://github.com/petrelharp/ftprime_ms/pull/37#discussion_r156617283
the simplification algorithm is at least linear in the number of edges of the input tree sequence.
Empirically the algorithm is exactly linear,
as seen in Figure~\ref{fig:simplify_complexity}B,
which shows the time required to simplify increasingly large subsets of a large tree sequence.
When simplifying the result of a forwards-time sequence, the number of edges is the main contributing factor.
Suppose on the other hand we want to
simplify an already-minimal but large tree sequence with $N$ nodes
to a subsample of size $n$.
How does the required time scale with $n$?
In this case, the computation is dominated by the size of the output tree sequence,
which grows with $\log(n)$, as shown in Figure~\ref{fig:simplify_complexity}C.

@@

In this article, we applied our methods for storing trees to the problem of pedigree recording in a forward-time
simulation.  However, the method applies to any simulation scheme generating nodes and edges.  For example, one could
use the methods described here to generate succinct tree sequences under coalescent processes not currently implemented
in \msprime{}, such as the coalescent with gene conversion \citep{Wiuf2000-rc}, using the structured coalescent to
model various forms of natural selection \citep{Kaplan1988-in,Kaplan1989-rt,Braverman1995-gn},
or the coalescent within a known pedigree.  For such models, one
could in principle generate tables of nodes and edges to be simplified in \tskit{}.  
The resulting succinct tree sequence object would be in the same format as those generated by \msprime{}'s simulate function,
and therefore compatible with existing methods for downstream analyses.
